### 2024-02-02 paper.tex
minor

This procedure is commonly adapted when working with single-cell DNA methylation data, because once one gets to the PCA step, one can then continue with the established methods just mentioned.
However, constructing a matrix suitable as input for PCA from methylation data requires deviation from the standard scRNA-seq workflow due to considerable differences in data structure.
First, while scRNA-seq quantifies RNA abundance of genes or transcripts, scBS is genome-wide and thus lacks a natural choice for features in which methylation is to be quantified.
Second, instead of counts, scBS generates binary data that informs us whether certain cytosines in a given cell are methylated or not.
A simple and common approach to construct a methylation matrix suitable for PCA, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g., 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (``methylation fractions'') between 0 and 1.
This matrix is now subjected to PCA.
After PCA, one can proceed with dimensionality reduction and clustering approaches known from scRNA-seq.

@@

To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Fig.~\ref{fig:score}E, and Methods for details). \new{The neighborhood relation in PCA space is relevant as it is used in many down-stream analyses, e.g., for clustering.}
We ask how the neighbor score depends on the number of available cells.
This is relevant as scBS protocols are costly and labor-intensive, and only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with all combinations of analysis methods, and calculated the mean neighbor score for each (Figure \ref{fig:score}D).
\new{
The results confirmed that quantifying methylation at VMRs leads to a cleaner separation of cell states than using promoter regions or 100~kb tiles.
Using the shrunken mean of residuals as a measure of DNA methylation improved results further.
This effect was most noticeable when using promoters or genomic tiles as intervals, presumably because an individual promoter region or tile might span genomic regions with varying levels of DNA methylation as depicted in Fig.~\ref{fig:smoothres}B, which our method accounts for.
Although cell cluster separation generally becomes more difficult in those, performance gains were observed also in smaller data sets.
}

@@

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.

@@

We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
\new{It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
However, one can also use to the same end established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarking]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
Once a low dimensional embedding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Figure \ref{fig:umaps}B).}

>>>>

This procedure is commonly adapted when working with single-cell DNA methylation data, because once one gets to the PCA step, one can then continue with the established methods just mentioned.
However, constructing a matrix suitable as input for PCA from methylation data requires deviation from the standard scRNA-seq workflow due to considerable differences in data structure.
First, while scRNA-seq quantifies RNA abundance of genes or transcripts, scBS is genome-wide and thus lacks a natural choice for features in which methylation is to be quantified.
Second, instead of counts, scBS generates binary data that informs us whether certain cytosines in a given cell are methylated or not.
A simple and common approach to construct a methylation matrix suitable for PCA, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g., 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Fig.~\ref{fig:smoothres}A).
This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (``methylation fractions'') between 0 and 1.
This matrix is now subjected to PCA.
After PCA, one can proceed with dimensionality reduction and clustering approaches known from scRNA-seq.

@@

To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Fig.~\ref{fig:score}E, and Methods for details). \new{The neighborhood relation in PCA space is relevant as it is used in many down-stream analyses, e.g., for clustering.}
We ask how the neighbor score depends on the number of available cells.
This is relevant as scBS protocols are costly and labor-intensive, and only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with all combinations of analysis methods, and calculated the mean neighbor score for each (Fig.~\ref{fig:score}D).
\new{
The results confirmed that quantifying methylation at VMRs leads to a cleaner separation of cell states than using promoter regions or 100~kb tiles.
Using the shrunken mean of residuals as a measure of DNA methylation improved results further.
This effect was most noticeable when using promoters or genomic tiles as intervals, presumably because an individual promoter region or tile might span genomic regions with varying levels of DNA methylation as depicted in Fig.~\ref{fig:smoothres}B, which our method accounts for.
Although cell cluster separation generally becomes more difficult in those, performance gains were observed also in smaller data sets.
}

@@

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Fig.~\ref{fig:workflow}).
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.

@@

We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
\new{It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
However, one can also use to the same end established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarking]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
Once a low dimensional embedding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Fig.~\ref{fig:umaps}B).}
